---
title: "Introduction to ontologyX"
author: "Daniel Greene"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ontologyX}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---
```{r echo=FALSE}
knitr::opts_chunk$set(dev="svg", fig.width=7, fig.height=5, dev="svg", fig.align="center")
```
`ontologyIndex` is the foundation of the 'ontologyX' packages: 

* `ontologyIndex`, for representing ontologies as R objects and enabling simple queries, 
* `ontologySimilarity`, for computing semantic similarity between ontological terms and annotations, 
* `ontologyPlot` for visualising sets of ontological terms with various graphical options. 

The functionality of the `ontologyIndex` package is centered around `ontology_index` objects: simple R representations of ontologies consisting of `list`s and `vector`s of term properties (ID, label, etc.) which are named by term so that simple look-ups by term can be performed. Ontologies encoded in OBO or OWL format can be read into R using the function `get_ontology`, or created explicitly using the function `ontology_index`. The minimum requirement for creating an `ontology_index` object is that each term should have an ID, a label and a set of parents/superclasses (possibly empty). The package comes with three such ready-made `ontology_index` objects: `hpo`, `mpo` and `go`, encapsulating the Human Phenotype Ontology (HPO), Mammalian Phenotype Ontology (MPO) and Gene Ontology (GO) respectively, each loadable with `data`. Here we'll demonstrate the package using the HPO.

```{r}
library(ontologyIndex)
data(hpo)
```

Note: to use an up-to-date version of a given ontology, download the relevant `.obo` or `.owl` file and read it into R using the function `get_ontology`:

```{r eval=FALSE}
ontology <- get_ontology(file)
```

The `ontology_index` object is just a list of 'vectors and lists' of term properties, indexed by the IDs of the terms:

```{r, echo=FALSE}
data.frame(property=names(hpo), class=sapply(hpo, class), stringsAsFactors=FALSE, row.names=NULL)
```

The 'special' term properties, which all `ontology_index` objects contain, are `id`, `name`, `parents`, `children` and `ancestors`. These are the properties which the functions in the `ontologyIndex` package operate on - however, additional properties per term - custom properties, or whichever properties are contained in the original file - can also be read in and queried in the same way (see the vignette 'Creating an ontology_index'). Custom functions which operate on additional properties can also be defined. The `children` and `ancestors` properties are based on propagating the `parent` relationships for each node. If reading an ontology from a file, the definition of the `parents` of each term can be defined as desired - for instance it could reflect `is_a`/'superclass' relationships, or `part_of` relationships, or both (see 'Creating an ontology_index' and `?get_ontology` for more details).

You can use the function `get_term_property` to query the `ontology_index` object, and retrieve a particular attribute for a single term. For instance:
```{r}
get_term_property(ontology=hpo, property="ancestors", term="HP:0001873", as_names=TRUE)
```

However you can also look up properties for a given term using `[` and `[[` as appropriate, since an `ontology_index` just a `list`. This is the best way to use the `ontology_index` if you are operating on multiple terms as it's faster.

```{r}
hpo$name["HP:0001873"]
hpo$id[grep(x=hpo$name, pattern="Thrombocytopenia")]
hpo$ancestors[["HP:0001873"]]
hpo$name[hpo$ancestors[["HP:0001873"]]]
```

## Removing redundant terms

A set of terms (i.e. a `character` vector of term IDs) may contain redundant terms. The function `minimal_set` removes such pairs so as to leave a *minimal set* of terms, in the sense of the ontology's directed acyclic graph.

```{r}
terms <- c("HP:0001871", "HP:0001873", "HP:0011877")
hpo$name[terms]
minimal <- minimal_set(hpo, terms)
hpo$name[minimal]
```

## Finding all ancestors of a set of terms

To find all the ancestors of a set of terms, i.e. all the terms which are an ancestor of any term in the given set, on can use the `get_ancestors` function:
```{r}
get_ancestors(hpo, c("HP:0001873", "HP:0011877"))
```

## Operating on subclasses 

There are functions which allow set operations with respect to descendancy: `intersection_with_descendants`, `exclude_descendants` and `prune_descendants`. Each function accepts a set of terms `terms` and a set of root terms `roots`. We here demonstrate how these functions work on a set of terms containing: 

* `"HP:0000707"`: Abnormality of the nervous system,
* `"HP:0000924"`: Abnormality of the skeletal system, 
* `"HP:0000006"`: Autosomal dominant inheritance.

The panels below show the original, untransformed set of terms, and the set of terms after each of the above mentioned functions has been called on it, passing `"HP:0000118"`: 'Phenotypic abnormality' as the `roots` argument (terms present in the set highlighted in green, those not present in grey). `ontologyPlot` was used to create the diagrams. 

```{r echo=FALSE}
library(ontologyPlot)
in_set <- c("HP:0000707", "HP:0000924", "HP:0000006")
show <- c(in_set, "HP:0000118", "HP:0000001")
fns <- list(
	`Original set`=function(ontology, roots, terms) terms, 
	`intersection_with_descendants of 'Phenotypic abnormality'`=intersection_with_descendants,
	`exclude_descendants of 'Phenotypic abnormality'`=exclude_descendants,
	`prune_descendants of 'Phenotypic abnormality'`=prune_descendants
)
par(mfrow=c(2,2))
for (i in seq_along(fns)) plot(cex.main=0.8, main=names(fns)[i], onto_plot(ontology=hpo, label=official_labels(hpo, show), terms=show, color=if (i == 1) "transparent" else ifelse(show == "HP:0000118", "red", "transparent"), fillcolor=ifelse(show %in% fns[[i]](hpo, roots="HP:0000118", terms=in_set), "#66FF66", "#DBE6E0"), fontsize=15))
par(mfrow=c(1,1))
```

For more details see the help file for the individual functions, e.g. `?exclude_descendants`. Note that to perform analagous operations with respect to sets of ancestors, one can use the `get_ancestors` function in conjunction with the base R set functions, e.g. `setdiff` and `intersect`.

## Additional ontological functionality

The packages `ontologySimilarity` and `ontologyPlot` can be used to calculate semantic similarity between and visualise terms and sets of terms respectively: see the corresponding vignettes for more details. 
