---
title: "Creating an `ontology_index`"
author: "Daniel Greene"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating an ontology_index}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

An `ontology_index` can be created by loading a pre-existing one - for example by calling `data(hpo)`, reading ontologies encoded in OBO or OWL format into R using `get_ontology`, or by calling the function `ontology_index` explicitly. An `ontology_index` is a named `list` of properties for each term, where each property is represented by a `list` or `vector` named by term (thus lookups by term name are simple). All `ontology_index` objects contain the `id`, `name`, `parents`, `children` and `ancestors` for each term, however additional properties can be added. For details on how to use an `ontology_index`, see the 'Introduction to ontologyX' vignette.

## Reading in from a file

The functions `get_OBO`, `get_OWL` and `get_ontology` can read ontologies into R as `ontology_index` objects, the latter determining which function to call based on the file name. By default, the properties `id`, `name`, `parents`, `children`, `ancestors` and `obsolete` are populated. 

*Note: the `xml2` package is required in order to read ontologies encoded in OWL format.*

The syntax is simple:
```{r eval=FALSE}
ontology <- get_ontology(file)
```

Additional information is often present in the original files - for example definitions (the `def` field in OBO format) and `part_of` relationships to other terms. It is possible to include these extra properties by passing additional arguments to `get_ontology` or `get_OBO`/`get_OWL` as appropriate. The given name of the additional arguments then go on to form the names of the properties included in the returned `ontology_index`, and the format of the arguments depends on whether you are reading an OBO or OWL formatted file. 

For OBO files, these arguments should be character vectors where the first element is a regular expression specifying what text to extract as the property and the second element giving the mode of the returned array (defaulting to `character`, but potentially also taking the values `logical`,`numeric`,`list`, etc.). Here for example, we additionally extract the `"definition"` field for each term:

```{r eval=FALSE}
ontology <- get_OBO(file, definition=c("^def: (\\.*)", "character"))
```

Now, `ontology` will have the `ontology$definition` field, a character vector (named by term as usual) populated, in addition to the standard properties. Thus, definitions can be extracted simply using `ontology$definition[termID]`.

For OWL files the additional arguments should be functions which given an `xml_nodeset` (see the `xml2` package), corresponding to all the terms in the ontology, return lists or vectors of term properties. A selection of such functions are included with the package, including `OWL_IDs`, `OWL_labels`, `OWL_obsolete` and `OWL_is_a`. The following expressions can be used to load `ontology_index`es with the `part_of` relationship populated from OWL formatted ontologies:

```{r eval=FALSE}
ontology <- get_OWL(file, part_of=OWL_part_of)
```

Here, `OWL_part_of` is a function which returns a list of `character` vectors containing the `part_of` relationships for each term. Customs functions based on XPath expressions can be defined if other term properties are desired, see `?OWL_strings_from_nodes` for details.

By default, if there are terms in parent-offspring relationships specified by the ontology which are missing, an error is thrown. However, the ontology reading functions `get_ontology`, `get_OWL` and `get_OBO` have a `remove_missing` argument (`FALSE` by default), which causes parent-offspring relationships with a missing term to be ignored, hence the complete parts of the ontology can still be used. 

When reading ontologies from OWL files, the term IDs are often given as URLs: 

`"http://purl.obolibrary.org/obo/<ontology code>_<number>"`, 

and thus the term IDs stored by the `ontology_index` object obtained with `get_OWL` will also be in this form. However, data is often supplied with the OBO ID format: `<ontology code>:<number>`. Possible solutions to making such data compatible with the ontologyX packages are to transform the term IDs in the data to the OWL format (i.e. to match the index) or transform the `ontology_index` object itself (to match the data, so that it's IDs are in OBO style). The former is simpler and can be done by `paste`ing data strings onto the appropriate URL and replacing the underscore with a colon. The latter can be done with the help of regular expressions, for example as demonstrated below, by looping over the properties in the `ontology_index`, and transforming them as appropriate depending on whether they are lists or vectors:

```{r eval=FALSE}
term_url <- "http://purl.obolibrary.org/obo/(\\w+)_(\\d+)"
for (property in names(ontology)) {
	if (is.character(unlist(use.names=FALSE, ontology[[property]]))) ontology[[property]] <- if (is.list(ontology[[property]])) lapply(ontology[[property]], function(term_properties) gsub(x=term_properties, pattern=term_url, replacement="\\1:\\2")) else gsub(x=ontology[[property]], pattern=term_url, replacement="\\1:\\2")
	names(ontology[[property]]) <- gsub(x=names(ontology[[property]]), pattern=term_url, replacement="\\1:\\2")
}
```

## Adding term properties

Modifying an existing `ontology_index` to add term properties is as simple as modifying a list. In the example below, we'll add the number of children for each term.

```{r eval=FALSE}
ontology$number_of_children <- sapply(ontology$children, length)
```

In the same manner, a valid `ontology_index` can be built up from scratch as a list, of course requiring that the standard properties are included for use with functions in `ontologyIndex`.


